---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ScrambledTreeBuilder

<!-- badges: start -->
<!-- badges: end -->

The goal of ScrambledTreeBuilder is to allow users to conveniently produce phylogenetic trees between species with the purpose of studying genome scrambling.

## Installation

You can install the development version of ScrambledTreeBuilder from [GitHub](https://github.com/) with:
      
``` {r}
# install.packages("devtools")
devtools::install_github("brennern/ScrambledTreeBuilder", repos=BiocManager::repositories()) |> suppressMessages()
```

## Usage
After performing an All Vs. All genome comparison between dozens of species, you may have .yaml files as the output. In order to convert these files into a dataframe in R, you may use the `formatStats()` function. Extract your .yaml files from their respective directory and store the information under the variable `yamlFiles`.

```{r}
library(ScrambledTreeBuilder) |> suppressPackageStartupMessages()

resultsDir <- system.file("extdata/yaml", package = "ScrambledTreeBuilder")
yamlFileData <- list.files(resultsDir, pattern = "*.yaml.bz2", full.names = TRUE)
names(yamlFileData) <- yamlFileData |> basename() |> sub(pat = ".yaml.bz2", rep="")

exDataFrame <- formatStats(yamlFileData)
```

To build the phylogenetic trees, your data frame will need to be transformed into a matrix. The function `makeMatrix()` will accomplish this.

```{r}
valuesToBuildTheTree <- "percent_identity_global"
treeMatrix <- makeMatrix(exDataFrame, valuesToBuildTheTree, 100, 50)
valuesToPlaceOnLabels <- "index_avg_strandRand"
valueMatrix <- makeMatrix(exDataFrame, valuesToPlaceOnLabels, 1, 0.5)
```

Then, in order to plot the percent identity and strand randomisation index scores on the tree, you will need to extract the tree data in a tibble and utilize the functions `makeValueTibble()`.

```{r}
HClust <- hclust(dist(treeMatrix), method = "complete")
Tibble <- tidytree::as_tibble(tidytree::as.phylo(HClust))
tibbleWithValue <- makeValueTibble(Tibble, valueMatrix, colname = "Strand_Randomisation_Index")
tibbleWithMultipleValues <- makeValueTibble(tibbleWithValue, treeMatrix, colname = "Percent_Identity")
```

Finally, to visualize your phylogenetic tree, you can utilize the `visualizeTree()` function by inputting your tibble data containing your desired variable.

```{r}
SingleValueTree <- visualizeTree(tibbleWithValue, tibbleWithValue$Strand_Randomisation_Index)

SingleValueTree +
  ggplot2::ggtitle(paste("Tree built with Percent Identity and labelled with Strand Randomisation Index scores")) + 
  viridis::scale_color_viridis(name = "Strand Randomisation Index") +
  ggtree::geom_hilight(node = 8, fill = "lightblue1", alpha = .2, type = "gradient", gradient.direction = 'tr') +
  ggtree::geom_hilight(node = 9, fill = "pink", alpha = .2, type = "gradient", gradient.direction = 'tr')
```
```{r}
MultiValueTree <- visualizeTree(tibbleWithMultipleValues, tibbleWithMultipleValues$Strand_Randomisation_Index, ynudge = 0.2)

MultiValueTree +
  ggplot2::ggtitle("Tree labeled with Strand Randomisation Index and Percent Identity (built with Percent Identity)") +
  viridis::scale_color_viridis(name = "Strand Randomisation Index") +
  ggnewscale::new_scale_colour() +
  ggtree::geom_label(ggtree::aes(label=round(Percent_Identity, digits = 3), color = Percent_Identity), label.size = 0.25, size = 3, na.rm = TRUE, label.padding = ggtree::unit(0.15, "lines"), nudge_y = -0.2) +
  viridis::scale_color_viridis(option = "magma", name = "Percent Identity")
```
