---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ScrambledTreeBuilder

<!-- badges: start -->
<!-- badges: end -->

The goal of ScrambledTreeBuilder is to allow users to conveniently produce phylogenetic trees between species with the purpose of studying genome scrambling.

## Installation

You can install the development version of ScrambledTreeBuilder from [GitHub](https://github.com/) with:
      
``` r
# install.packages("devtools")
devtools::install_github("brennern/ScrambledTreeBuilder")
```

## Usage
After performing an All Vs. All genome comparison between dozens of species, you may have .yaml files as the output. In order to convert these files into a dataframe in R, you may use the `formatStats()` function. Extract your .yaml files from their respective directory and store the information under the variable `yamlFiles`.

```{r}
library(ScrambledTreeBuilder)

resultsDir <- system.file("extdata/PairwiseComparisons", package = "ScrambledTreeBuilder")
yamlFileData <- list.files(resultsDir, pattern = "*.yaml.bz2", full.names = TRUE)
names(yamlFileData) <- yamlFileData |> basename() |> sub(pat = ".yaml.bz2", rep="")

exDataFrame <- formatStats(yamlFileData)
```

To build the phylogenetic trees, your data frame will need to be tranformed into a matrix. The function `makeMatrix()` will accomplish this.

```{r}
valuesToBuildTheTree <- "percent_identity_global"
treeMatrix <- makeMatrix(exDataFrame, valuesToBuildTheTree, 100, 50)
valuesToPlaceOnLabels <- "index_avg_strandRand"
valueMatrix <- makeMatrix(exDataFrame, valuesToPlaceOnLabels, 1, 0.5)
```

Then, in order to plot the percent identity and strand randomisation index scores on the tree, you will need to extract the tree data in a tibble and utilize the functions `makeValueTibble()`.

```{r}
HClust <- hclust(dist(treeMatrix), method = "complete")
Tibble <- tidytree::as_tibble(tidytree::as.phylo(HClust))
tibbleWithValue <- makeValueTibble(Tibble, valueMatrix)
```

Finally, to visualize your phylogenetic tree, you can utilize the `visualizeTree()` function by inputting your tibble data containing your desired variable.

```{r}
Tree <- visualizeTree(tibbleWithValue, tibbleWithValue$value)

Tree + 
  ggplot2::ggtitle(paste("Tree built with", valuesToBuildTheTree, "and labelled with", valuesToPlaceOnLabels)) + 
  viridis::scale_color_viridis(name = valuesToPlaceOnLabels)
```
