---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ScrambledTreeBuilder

<!-- badges: start -->
<!-- badges: end -->

The goal of ScrambledTreeBuilder is to allow users to conveniently produce phylogenetic trees between species with the purpose of studying genome scrambling.

## Installation

You can install the development version of ScrambledTreeBuilder from [GitHub](https://github.com/) with:
      
``` r
# install.packages("devtools")
devtools::install_github("brennern/ScrambledTreeBuilder")
```

## Usage
After performing an All Vs. All genome comparison between dozens of species, you may have .yaml files as the output. In order to convert these files into a dataframe in R, you may use the `ScrambledTreeBuilder::formatStats()` function. Extract your .yaml files from their respective directory and store the information under the variable `yamlFiles`.

```{r}
library(ScrambledTreeBuilder)

params <- list()
params$resultsDir <- '/flash/LuscombeU/noa/halobacteria/results/genomicbreaks'
yamlFiles <- list.files(params$resultsDir, pattern = "*.yaml", full.names = TRUE)
names(yamlFiles) <- yamlFiles |> basename() |> sub(pat = ".yaml", rep="")

dataframe <- formatStats()
```

To build the phylogenetic trees, your data frame will need to be tranformed into a matrix. The function `ScrambledTreeBuilder::makeMatrix()` will accomplish this.

```{r}
m <- makeMatrix(dataframe, "percent_identity_global", 100, 50)
hclust <- hclust(dist(m), method = "complete")
dend <- as.dendrogram(hclust)

m1 <- makeMatrix(dataframe, "index_avg_strandRand", 100, 50)
```

Then, in order to plot the percent identity and strand randomisation index scores on the tree, you will need to extract the tree data in a tibble and utilize the functions `ScrambledTreeBuilder::extractPercentID()` and `ScrambledTreeBuilder::extractStrandRandIDs()`.

```{r}
as_tibble(as.phylo(dend)) -> td
sRIs <- unique(td$parent) |> sort() |> purrr::set_names() |> sapply(extractStrandRandIDs, td, m1)
pIDs <- unique(td$parent) |> sort() |> purrr::set_names() |> sapply(extractPercentID, td, m)
td$strandRand <- NA
td$percentID <- NA
td[names(sRIs), "strandRand"] <- unname(sRIs)
td[names(pIDs), "percentID"] <- unname(pIDs)
```
